Software Development in Startup Companies: The Green-field Startup Model

### 2013-05-13 article/main.tex
ready for tony review

In this section we present relevant implications which emerge from the behavior
of early-stage startups, formally expressed in the GSM, answering to our
research question (\textit{What is the current state-of-practice related to
software development strategy in early-stage startups?}) through its 
sub-questions. Although the startups we inquired in the case study were spread
across different nations and market sectors, we have found several patterns
that emerged very clearly.

@@

The most urgent priority of software development in startups is to shorten
time-to-market. When developing a new product, which has not been attempted
before, it is crucial for the company to release the product soon and observe
the users reactions. The company's chances to survive, are strictly related to
the ability to find the right product/market fit as quickly as possible.
Developing software without releasing it for a long period of time, coming up
with very complex solutions is often not a viable option. Business demands,
investors interests, market pressures and the general uncertain conditions force
startups to quickly iterate on the product by exploring and experimenting new
functionalities progressively rolled-out to early-adopters. 
%MUN1: Ok, I'm now "inventing" possible implications in the sections below...  
%this is your job and you are much more qualified to do that than me :-)

@@

%------------------------------------------------------------------------------%
\subsection{Light-weight methodology} 
Startups do not apply any standard development methodology. When focused on 
building the first version of the product, startups do not observe any 
specific and standard development methodologies or processes. From an accurate 
and in-depth investigation into daily activities, it emerged that the number of 
the observed development \textit{practices} is extremely small and never 
systematic (only little \textit{pair programming} sessions and rarely 
\textit{test driven development}). Despite the wide number of publications, 
which discuss the adoption of lightweight methodologies, we have found that 
especially in this first stage not only methodologies but even individual 
practices are basically neglected.

@@

%MUN1: paragraph below is candidate for removal The greatest part of engineering
%activities of startups are focused on  the implementation while only little
%attention is given to more conventional  activities (project management,
%requirement specifications, analysis,  architecture design, automatic testing,
%\ldots). Especially when bringing the  first product to the market, the
%attitude of startups towards traditional  engineering activities is strongly
%characterized by a general feeling of  unnecessary. Every document, artifact or
%process, which does not produce a  short-term tangible benefit to the software
%production is perceived as a waste  of time. CG: ok

%------------------------------------------------------------------------------%
\subsection{Focus on minimal set of functionalities} 
The first release of the product includes only a limited set of well suitable 
functionalities focused on user experience. These functionalities are 
usually set by the founders and their experience in a particular business 
domain. As soon as possible they provide demos to demonstrate the idea to 
potential customers. 
%MUN1: This rises the question, how do startups prioritize requirements.  
%There's abundant amount of literature on how to select/prioritize reqs., but  
%do they work in startups? Are they also too heavyweight or could they be  
%adopted by startups in a modified form in order to increase the probabilities  
%of success in the market?. CG: fixed. 
As confirmed by practitioners, the user must be able to accomplish a task on 
the product without major interruptions in the flow, which would cause 
frustration before being able to capture his attention. To realize UX in a 
short time frame, developers focus on a limited number of suitable 
functionalities, especially in the first version(s) of the product. Other 
product qualities are regarded as less important: for instance, obtaining a 
high-reliability is not a priority since early adopters of innovative product 
are keen to accept minor bugs in exchange of new functionalities\footnote{These 
results further highlight differences between software development in startups 
and traditional web engineering, where quality aspects are rated as very 
important, even more than time-to-market~\cite{Offutt2002}.}. From a business 
perspective, one might argue that established companies might take advantage of 
the new idea and develop a better version of the product internally. Despite 
this might represent a risk, established companies usually cannot afford to 
operate in a market of \textit{innovators} and \textit{early adopters} group 
of users - described by the Roger's bell curve in~\cite{Rogers} - because of 
the little return of investment they would gain, as discussed by Christensen 
in~\cite{Christensen2003}. Instead, established companies have more interests 
in acquiring a startup and all the knowledge it has gained during the 
\textit{customer development} process while focusing effort on acquiring larger
shares of a ``mass market'' sector. 
%MUN1: candidate for shorting 
%Furthermore, thanks to cloud computing and more in general high availability 
%of third party components, aspects such as performance, security and 
%scalability can be achieved with a reasonable effort. Thus, given the 
%general lack of quality concerns, and supported by the fact that the optimal 
%strategy to improve the UX is by having the product used by real customers, 
%startups can effectively focus on achieving short time-to-market. CG: fixed

@@

%------------------------------------------------------------------------------%
\subsection{Paying back the technical debt} 
The initial lack of structures and processes negatively affects the 
performance when the company starts to grow.
One of the primary objectives of startups is to expand their business into fast
growing markets. If successful, the company will face a growth in terms of
number of customers, employees and product capabilities. Then, the increased
complexity arises the necessity of controlling the initially chaotic software
development environment. Under delicate circumstances, where an increased
market demand meets a poorly engineered product and process, the development
team need to start returning the accumulated technical debt. As reported by
practitioners, introducing more structures, documentation and standard
procedures, cause an initial drop-down in performance which severity depends on
the speed of growth, the amount of technical debt and the capacity of the team
to absorb it. 
%MUN1: what does literature say about technical debt? Are there strategies that 
 %maybe applicable to startups and if not why? Is there more research needed?  
%Obviously, shortening time-to-market and preventing technical debt are two  
%conflicting goals, but is there a chance to find a sweet spot where  
%development speed can be balanced to achieve market success and still reducing 
%the performance bump that hampers growth? 
%CG: partially fixed below in summary.

>>>>

In this section we present relevant implications which emerge from the behavior
of early-stage startups, formally expressed in the GSM, answering to our
research question (\textit{What is the current state-of-practice related to
software development strategy in early-stage startups?}) through its 
sub-questions. Although the startups we inquired in the case study were spread
across different nations and market sectors, we have found some patterns
that emerged very clearly.

@@

The most urgent priority of software development in startups is to shorten
time-to-market. When developing a new product, which has not been attempted
before, it is crucial for the company to release the product soon and observe
the users reactions. The company's chances to survive, are strictly related to
the ability to find the right product/market fit as quickly as possible.
Developing software without releasing it for a long period of time, coming up
with very complex solutions is often not a viable option. Business demands,
investors interests, market pressures and the general uncertain conditions force
startups to quickly iterate on the product by exploring and experimenting new
functionalities progressively rolled-out to early-adopters.

Introducing a software development process in this early phases of high pressure can be challenging for startups. However, some development approaches, mostly investigated in mature companies, could help in shortening time-to-market. Among several methodologies that can be found in literature, the rapid application development (RAD) \cite{mcconnell2004professional} facilitates prototyping and promotes a continuous interactive process that allows early-adopters to understand, modify and approve a working product that meets their needs. A study of Gerbel et. al. states that the use of RAD, which promotes minimal planning and rapid prototyping, allows software to be released faster and makes easier to change requirements \cite{Gerber_practicalimplications}. 
%MUN1: Ok, I'm now "inventing" possible implications in the sections below...  
%this is your job and you are much more qualified to do that than me :-)

@@

%------------------------------------------------------------------------------%
\subsection{Light-weight methodology} 
Startups do not apply any standard development methodology. When focused on 
building the first version of the product, startups do not observe any 
specific and standard development methodologies or processes. From an accurate 
and in-depth investigation into daily activities, it emerged that the number of 
the observed development \textit{practices} is extremely small and never 
systematic (only little \textit{pair programming} sessions and rarely 
\textit{test driven development}).

@@

While the absence of a basic process might enable startups to focus more on the product, some studies claim that companies can take advantage from the use of some engineering activities even in the early stages \cite{989006}. For instance an experience report conducted by Taipale \cite{Taipale2010} shows how their startup took advantage from tailoring some simple XP practices to their needs. 
%MUN1: paragraph below is candidate for removal The greatest part of engineering
%activities of startups are focused on  the implementation while only little
%attention is given to more conventional  activities (project management,
%requirement specifications, analysis,  architecture design, automatic testing,
%\ldots). Especially when bringing the  first product to the market, the
%attitude of startups towards traditional  engineering activities is strongly
%characterized by a general feeling of  unnecessary. Every document, artifact or
%process, which does not produce a  short-term tangible benefit to the software
%production is perceived as a waste  of time. CG: ok

%------------------------------------------------------------------------------%
\subsection{Focus on minimal set of functionalities} 
The first release of the product includes only a limited set of well suitable 
functionalities focused on user experience. These functionalities are 
usually set by the founders and their experience in a particular business 
domain. As soon as possible they provide demos to demonstrate the idea to 
potential customers.

To ship a product with the right features built in, startups need to prioritize and filter. From an engineering point of view, most startups do not explicitly apply traditional Requirement Engineering (RE) activities to collect and manage requirements. However, some studies have found that by integrating simple techniques, such as Persona and Scenario, companies can improve the effectiveness of the requirements elicitation even whether the final users is mostly unknown \cite{1531030}. Another study suggest that using a lightweight project-initiation framework such as the Agile Inception Deck can help in preventing premature failure of the software project due to wrong understanding of the project requirements \cite{1667597}.
%MUN1: This rises the question, how do startups prioritize requirements.  
%There's abundant amount of literature on how to select/prioritize reqs., but  
%do they work in startups? Are they also too heavyweight or could they be  
%adopted by startups in a modified form in order to increase the probabilities  
%of success in the market?. CG: fixed. 
As confirmed by practitioners, the user must be able to accomplish a task on 
the product without major interruptions in the flow, which would cause 
frustration before being able to capture his attention. To realize UX in a 
short time frame, developers focus on a limited number of suitable 
functionalities, especially in the first version(s) of the product. Other 
product qualities are regarded as less important: for instance, obtaining a 
high-reliability is not a priority since early adopters of innovative product 
are keen to accept minor bugs in exchange of new functionalities\footnote{These 
results further highlight differences between software development in startups 
and traditional web engineering, where quality aspects are rated as very 
important, even more than time-to-market~\cite{Offutt2002}.}. From a business 
perspective, one might argue that established companies might take advantage of 
the new idea and develop a better version of the product internally. Despite 
this might represent a risk, established companies usually cannot afford to 
operate in a market of \textit{innovators} and \textit{early adopters} group 
of users - described by the Roger's bell curve in~\cite{Rogers} - because of 
the little return of investment they would gain, as discussed by Christensen 
in~\cite{Christensen2003}. Instead, established companies have more interests 
in acquiring a startup and all the knowledge it has gained during the 
\textit{customer development} process while focusing effort on acquiring larger
shares of a ``mass market'' sector. 
%MUN1: candidate for shorting 
%Furthermore, thanks to cloud computing and more in general high availability 
%of third party components, aspects such as performance, security and 
%scalability can be achieved with a reasonable effort. Thus, given the 
%general lack of quality concerns, and supported by the fact that the optimal 
%strategy to improve the UX is by having the product used by real customers, 
%startups can effectively focus on achieving short time-to-market. CG: fixed

@@

%------------------------------------------------------------------------------%
\subsection{Paying back the technical debt} 
The initial lack of structures and processes negatively affects the 
performance when the company starts to grow.
One of the primary objectives of startups is to expand their business into fast
growing markets. If successful, the company will face a growth in terms of
number of customers, employees and product capabilities. Then, the increased complexity requires the team to start controlling the initial chaotic software development environment. Under delicate circumstances, where an increased
market demand meets a poorly engineered product and process, the development
team need to start returning the accumulated technical debt. As reported by
practitioners, introducing more structures, documentation and standard
procedures, cause an initial drop-down in performance which severity depends on
the speed of growth, the amount of technical debt and the capacity of the team
to absorb it.

As stated by many interviewee, modern coding platforms (e.g. github\footnote{http://www.github.com}) and the use of well-known frameworks can be leveraged to mitigate the technical debt. Coding platforms allow developers to integrate in one place several engineering activities such as requirements lists, issue tracking, code repository, documentation, continuous integration, release and configuration management. Frameworks include support programs, compilers, code libraries and tool sets to enable the initial development of a project with limited overhead.

%MUN1: what does literature say about technical debt? Are there strategies that 
 %maybe applicable to startups and if not why? Is there more research needed?  
%Obviously, shortening time-to-market and preventing technical debt are two  
%conflicting goals, but is there a chance to find a sweet spot where  
%development speed can be balanced to achieve market success and still reducing 
%the performance bump that hampers growth? 
%CG: partially fixed below in summary.
